---
title: "Part 5 Data visualization"
author: "Savins"
date: "6/14/2022"
output: html_document
---

# Data visualization 
- Variation:within one variable
#a(continousvariable);Boxplot,histogram 
#b(categoricalvariable);pie chart,bar chart

- Co variation:Between two variables
#a)continous vs continous;scatter plots
#b)continous vs categorical;Box plot,mean bar graph
#c)categorical vs categorical;Stacked bar graph and clustered bar graph


- Categorical variables:classified variables with more than one group
- Continuous variable:variables that assume numerical values

## Library packages
```{r}
library(ggplot2)
library(dplyr)
library(tidyverse)
```

# importing and omiting NA's in the data
```{r}
combined_data<-read.csv('combined.csv',na.strings="")
combined_data1
combined_data1<-read.csv('combined.csv')
newCombined_data1<-combined_data1 %>% drop_na()

combined_data1 %>% drop_na()
combined_data %>% drop_na()

glimpse(combined_data)
View(combined_data)
colSums(is.na(combined_data))
view(combined_data)
newCombined<-combined_data %>% drop_na()
glimpse(newCombined)
colSums(is.na(newCombined))
View(newCombined)
newCombined

```

# checking and removing outliers
```{r}
#checking outliers
boxplot(newCombined$hindfoot_length)
#removing outliers
newCombined[!newCombined %in% boxplot.stats(newCombined$hindfoot_length)$out ]

```

#checking and removing duplicates
```{r}
#checking for duplicates
duplicated(newCombined)
sum(duplicated(newCombined))

  #removing duplicates
distinct(newCombined)
```

#removing empty rows and columns
```{r}
combined.data<-read.csv('combined.csv')
install.packages('janitor')
library(janitor)
remove_missing(combined.data)
clean_names(combined.data)
clean_names(combined.data,)
?clean_names()
```


# making plots
we use the ggplot() under the ggplot2
```{r}
#geoms- used for graphical representation of the data can be lines, points, bars
#geom_point() is used for scatter plot
#geom_lines() used trend lines for time series
#geom_bar() used for box plots
#geom_jitter gives the color
ggplot(data = newCombined,mapping = aes(x=weight,y=hindfoot_length))+geom_point()
ggplot(data = newCombined,mapping = aes(x=weight,y=hindfoot_length))+geom_point()+geom_jitter(colour='blue')

```

#hexbin
for a data set with large data observations eg combined dataset, over plotting of points can be a limitation of the scatter plot one strategy of handling such a setting is to use hexagonal binning of observations. The plot space is tessellated into hexagons. Each hexagon is assigned a color based on the number of observations that fall within its boundaries
```{r}
library(hexbin)
str(newCombined)
newCombined<-newCombined %>%  mutate(species=as.character(species))
ggplot(data = newCombined,mapping = aes(x=weight,y=hindfoot_length))+geom_boxplot()
plot(newCombined$weight,newCombined$hindfoot_length)
```

# alpha adds some transparency to your plot
```{r}
ggplot(data = newCombined,mapping = aes(x=weight,y=hindfoot_length))+
 geom_point(alpha=0.5,aes(colour=sex))
```

# Assignment
```{r}
View(newCombined)
str(newCombined)
fix(newCombined.m)
newCombined$sex.f<-as.factor(newCombined$sex)
n_distinct(newCombined$sex)
```
# alternative way of replacing data in your data set
```{r}
newCombined$sex<-car::recode(newCombined$sex,"'male'='Male';'Male'='Male';'M'='Male';'Female'='Female';'female'='Female';'F'='Female';'MALE'='Male';'FEMALE'='Female'")
```

# making a box plot
```{r}
newCombined$sex.f<-as.factor(newCombined$sex)
#making a box plot
ggplot(data = newCombined,mapping = aes(species_id,weight))+geom_boxplot()+theme_bw()

```


# Replacing values 
```{r}
newCombined.m %>% mutate(gender=ifelse(sex%in%c("MALE","M","m","male"),"Male","Female"))
```


# Plotting time series
Time series data is a set of observations on the values that a variable takes at different times
Cross-sectional data-This is the data of one or more variable collected at the same point in time
Lets calculate the no.of counts per year for each genres.First we need to group the data and count records within each group
```{r}
Yearly_count<-newCombined %>% count(year,genus)

Yearly_count
```

```{r}
Yearly_count1<-newCombined %>% group_by(year,genus) %>% summarise(n=n()) 
Yearly_count1
```

```{r}
ggplot(Yearly_count1,aes(year,n,color=genus))+geom_line()+facet_wrap(facets = vars(genus))

```
# assignment
```{r}
Yearly_count2<-newCombined %>% group_by(year,genus,sex) %>% summarise(n=n()) %>% filter(genus%in%c("Chaetodipus","Dipodomys","Reithrodontomys","Peromyscus"))
Yearly_count2
```

```{r}
ggplot(Yearly_count2,aes(year,n,color=genus))+geom_line()+facet_wrap(facets = vars(genus))+geom_point()

```

```{r}
ggplot(Yearly_count2,aes(year,n,color=sex))+geom_line()+facet_wrap(facets = vars(genus))+theme_bw()+labs(x='Year\n',y='Count\n',title="Sex Aginst Genus\n") +
    theme(axis.text.x = element_text(angle = 90),axis.title= element_text(size = 22))
```

# Converts blanks into NA
```{r}
combined_data[combined_data1=='']#How to check for blanks
combined_data1[combined_data1=='']<-NA
combined_data1 %>% na_if('')
combined_data1 %>% mutate_all(na_if,'')


```






























