---
title: "Part 2.Regression"
output: html_notebook
---
```{r}
install.packages("rcompanion")
library(rcompanion)
```

#converting non-linear data to linear working with their lamda's
```{r}
#we use the rcompanion package
transformTukey(auto_df$weight,plotit = FALSE,returnLambda = TRUE)
transformTukey(auto_df$mpg,plotit = FALSE,returnLambda = TRUE)
transformTukey(auto_df$year,plotit = FALSE,returnLambda = TRUE)
?transformTukey
auto_df$year2=auto_df$year^0.475
auto_df$mpg2=auto_df$mpg^0.175
auto_df$weight2=-1*auto_df$weight^-0.475

attach(auto_df)
auto_df

```

#splitting
```{r message=FALSE, warning=FALSE}
require(caTools)
sample = sample.split(auto_df,SplitRatio = 0.80) 
train1 =subset(auto_df,sample ==TRUE) # creates a training dataset named train1 with rows which are marked as TRUE
test1=subset(auto_df, sample==FALSE)
nrow(train1)
nrow(test1)
nrow(auto_df)
```
#final model
```{r}
fit5 <- lm(mpg2~weight2+year,data = train1)
fit6 <- lm(mpg~weight+year,data = train1)
summary(fit5)
#summary(fit6)
plot(fit5,which = 1)
#plot(fit6,which = 1)

```

## Accuracy

We will use the test data to check the accuracy of our final model.By considering the following ;
#1.R
```{r message=FALSE, warning=FALSE}
predictions<-predict(fit5, test1)
summary(fit5)
mse <- mean((test1$mpg - predictions)^2)
print(mse)
```


#.Error rate of the model
```{r message=FALSE, warning=FALSE}
(sigma(fit5)/mean(test1$mpg))*100

```


```{r message=FALSE, warning=FALSE}
test1$predicted<- predict(fit5,test1)
actuals_preds <- data.frame(test1$mpg,test1$predicted)
names(actuals_preds)<- c("mpg","predicted")
correlation_accuracy <- cor(actuals_preds)
correlation_accuracy
```
The correlation of the models shows that the accuracy is around 92 %.Lets check the true mpg and predicted mpg for the test data.

```{r}
head(actuals_preds)

```
Wow! what happened?We used log to the response value in our model which predicted the response with a log. If we use  log to the original mpg we can relate this result.

```{r message=FALSE, warning=FALSE}
actuals_preds$log_mpg <- log(actuals_preds$mpg)
head(actuals_preds)
test_dummy<-actuals_preds %>% select(-mpg)
cor(test_dummy)
 cor(actuals_preds$predicted,actuals_preds$log_mpg)
```

