---
title: "continuation on data science"
author: "Savins"
date: "5/18/2022"
output:
  word_document: default
  html_document: default
---
## lecture 2
The sav is a spss file
```{r}
library(ggplot2)
diamond<-diamonds[1:20,]
```

```{r}
#install.packages("haven")
#import spss data-sav
library(haven)
write_sav(diamond,"diamond6.sav")
write_dta(diamond,"diamond7.dta")

```

Importing data to R
```{r warning=FALSE}
```


```{r warning=FALSE}
data<-read.csv("diamond2.csv",header = TRUE,na.strings = '.')
```


```{r warning=FALSE}
data1<-read.csv("diamond2.csv",header = TRUE,na.strings = '.')
data2<-read.csv("diamond2.csv",header = TRUE,na.strings = '.')
data3<-read.csv("diamond2.csv",header = TRUE,na.strings = '.')

data
read.csv('diamond2.csv')
#The difference with the underscore and the dot is that we used a package called readr
#The readr package converts the data to a tibble rather than a data frame which is more sufficient and faster

read_csv('diamond2.csv')

read_sav("diamond6.sav")
```


```{r}
library(tidyverse)
```

## Identifying and Removing NA in you data
```{r}
#1
colSums(is.na(data))
#2
library(dplyr)
data_p<-data %>% drop_na()
```

## Replacing NA in your data with mean,median and random numbers
```{r}
#1
data[is.na(data)]<-mean(data$y)
data
#2
data1[is.na(data1)]<-median(data$y)
#3
data2[is.na(data2)]<-59
data3
```


```{r}
ma<-c(2,21,2,3,5,6,3,3,5,6,6,NA,2,3,4,5,6,6)
mean(ma,na.rm = TRUE)
```

## Replacing NA in categorical data

Mode imputation (or mode substitution) replaces missing values of a categorical variable by the mode of non-missing cases of that variable.

## Filtering Columns
tidyverse package contains both 
dplyr and tidyr
```{r}
library(tidyverse)
?iris
  iris %>%select(Sepal.Length,Species)
```

## dplyr package
dplyr package-makes data manipulation easy
Select()-is used to subset some columns
mutate()-is used to create new variables from existing variables
transmute()-
filter()-it subsets rows on condition
distinct()-gives unique cases in your variable/data
arrange()-orders the row
Rename()-changes the name of the variable
slice()-gives you a specific no. of rows
grepl()
```{r}
#select
mn<-c(rep(NA,times=50))
sum(mn)
is.na(iris2)

iris_b[(1:54)]<-c(rep(NA,times=54))
iris
iris<-cbind(iris2,iris_b)
iris2<-select(iris2,-iris_b)
iris<-iris %>% select(-Petal.Width)
iris<-iris %>% rename(.,Petal.Width=iris_b)
select(iris,Species,Sepal.Length,Sepal.Width)
names(iris)
iris %>% select(Species,Sepal.Length,Sepal.Width)
#ctrl+shift+m
iris %>% select(-(Species))
iris %>% select(-(Sepal.Length:Petal.Length))
iris %>% select(starts_with("p"))
iris %>% select_if(is.numeric)
#filter
iris %>% filter(Sepal.Length==5)
iris %>% filter(Species=='virginica',Sepal.Width==3)
iris %>% filter(Species=='virginica'|Sepal.Width==3)
iris %>% filter(Species=='virginica'&Sepal.Width==3)
iris %>% filter_all(any_vars(.>6.0))

#grepl
grepl()
grepl()
```

## Challenge 1

```{r}
library(ggplot2movies)
library(tidyverse)

movies
names(movies)
#1
movies %>% select(title:votes,Action:Short)
#2
movies %>% select(-(starts_with("r")),-(starts_with("m")))
#3
movies %>%
select_if(~sum(is.na(.))/length(.)>.25)
```


## revision on #1
```{r include=FALSE}
#1.To loop over each column
#2.Percentile
#3.compare the percentile to the data columns-condition
```
```{r}
```


```{r}
new_data<-mydata %>% mutate(biomas=height/weight)
new_data
new_data$biomas[c(2,4,6)]<-c(NA,NA,NA)
new_data$weight[c(1,3)]<-c(NA,NA)
new_data
perc<-function(x){
  for (i in 1:length(x)) {
   m<-as.list(((sum(length(which(is.na(x))))/length(x))*100)>30)
  }
return(m)
}
iris
k<-perc(iris)
k
perc(movies)
movies %>% select_if(~any(is.na(.)&perc(.)==TRUE))
iris %>% select_if(~any(is.na(.))&perc(.)==TRUE)

```



```{r eval=FALSE, include=FALSE}

new_data %>% select_if(~any(is.na(.))&n_distinct(.))
new
manigga<-as_mapper(~any(is.na(.))&n_distinct(.))
manigga(colSums(is.na(new_data)))
new_data %>% select_if(~any(is.na(.))) %>% sapply(., function(x) sum(is.na (x)))
```
## Revision
```{r}
new_data
movies %>%
    summarise(across(everything(), ~ sum(is.na(.))>5000)) %>% select_if(.,~isTRUE(.))

?napredict()
 sapply(z, function(x) sum(is.na (x)))
 new_data %>% select_if(~any(is.na(.))) %>% sapply(., function(x) sum(is.na (x)))
 iris %>%
filter_all(any_vars(abs(. - mean(.))>2*sd(.)))
```


## Challenge2

```{r}
#1
movies %>% filter(Action==1)
#2
movies %>% filter(length>=360)
names(movies)
#3
movies %>% filter_if(~is.integer(.)&n_distinct(.)<3, all_vars(.<1))
movies %>% filter_at(vars(Action:Short), all_vars(.==0))

```
## Renaming data columns
```{r}
#Changing the names
iris %>%
select_all(str_to_lower)
iris %>%
rename(sw=Sepal.Width,sl=Sepal.Length)
#
iris %>%
rename_at(vars(starts_with("S")), str_to_lower)

iris %>%
rename_if(is.numeric, str_to_lower)
```

## a) Mutate()
```{r}
iris %>% mutate(biomas=Sepal.Length/Sepal.Width)
iris %>% transmute(biomas=Sepal.Length/Sepal.Width)
iris %>% slice(1:5)
iris %>% slice(-(1:floor(n()/3)))
iris
?floor

?n
```
## b)
```{r}
mtcars
names(mtcars)
mtcars<-mtcars %>%
rownames_to_column("car")
```
```{r}
diamond2
names(diamond2)
diamond2 %>% rename(index=X)
rownames_to_column("Index")
mydataset
```


## revision
```{r}
#new_data %>%  table(is.na(.))
new_data
table(is.na(new_data))
 new_data1<-new_data %>% rownames_to_column("index")
n
#1
function(x){
for (i in 1:col(x)) {
 as.list((colSums(is.na(x))/nrow(x)*100)>52)

}
}
new_data %>% select_if(~any(is.na(.))) %>% sapply(.,function(x)sum(length(which(is.na(x))))/length(x))



new_data %>% select_if(~any(is.na(.))) %>% sapply(., function(x) sum(is.na (x)))

new_data %>% select_if(~any(colSums(is.na(.))/nrow(.)*100)>52)
new_data %>% select_if(~colSums(is.na(.==TRUE)))
is.numeric(new)


```










```{r}
library(ggplot2movies)
library(tidyverse)
movies %>% rename(.,budget_if_known=budget)
movies %>% select_if(~sum(is.na(.))/length(.)>.25,str_to_upper)
```
## inclusion operator
```{r}

#inclusion operator->%in%
movies %>% filter(title%in%c("$21 a Day Once a Month","$40,000","'Hukkunud Alpinisti' hotell"))
movies
```

## Filtering rows
```{r}
movies %>% filter(title=="1 Giant Leap"&year==2002)
movies %>% filter(year>2004)
movies %>% filter(rating >6.5,title%in%c("$21 a Day Once a Month","$40,000","'Hukkunud Alpinisti' hotell"))

?contains
```


```{r}
iris %>%
filter_all(any_vars(.>7.5))
iris %>%select(-Species) %>% 
filter_all(any_vars(abs(. - mean(.))>2*sd(.)))
iris %>% filter_if(is.numeric,all_vars(.<mean(.)))

```


Re-arranging your data
# sorting rows
By default, arrange() sorts columns in ascending order i.e. from smallest to largest.
To change the sort order to descending for a column you put the function desc() around the column name.

```{r}
iris %>%
arrange_if(~is.factor(.), desc)
str(iris)
iris
```
## Re-ordering columns
You can use the select() function to re-order columns.
```{r}
iris %>%
select(starts_with("P"), everything())

#To sort alphabetically, we need to extract column names which we can do using the function current_vars() and then sort the names

iris %>%
select(sort(current_vars()))
```


## challenge 3
```{r}
movies %>% arrange(desc(title))
movies %>% select_if(~n_distinct(.)==2)  
movies
fix(iris)
iris
```

## changing data
## mutate
-This section will focus on how you can update, drop, and create new columns
-To make changes, we mutate() our dataset.
-The functions we use in the expressions need to work on vectors
and operate just like how they would behave if we were working on
the vector independently of the data.frame.1 If we put a name on
the LHS of an expression and an = between them, we give the new column a name
-toupper() converts the string to upper case or lower case and vice versa

```{r}
iris %>%
mutate(Sepal.Area = Sepal.Width * Sepal.Length)
#When using mutate() you can create multiple new columns

iris %>%
mutate(Sepal.Area = Sepal.Width * Sepal.Length,
Avg.Sepal.Area = mean(Sepal.Area))

#Updating a column involves using its name on the LHS of an assign.

iris %>%
mutate(Sepal.Width=Sepal.Width*.9)

#To delete a column:
# 1)you can either assign NULL
iris %>% mutate(Sepal.Length=NULL)

#2) use the exclude syntax for a select()
iris %>% select(-Sepal.Length)

#In general you can perform multiple task at once enabling readability
iris %>%
mutate(Sepal.Area = Sepal.Width * Sepal.Length,
Avg.Sepal.Area = mean(Sepal.Area),
Sepal.Width=Sepal.Width*.9,
Sepal.Length=NULL)
```

```{r}
#We can use row_number() to add an ID column to our data.
iris %>% mutate(id=row_number())

#lag() $ lead()
iris %>%
mutate(prev = lag(Sepal.Length),
future = lead(Sepal.Length))

?lag
x<-1:5
tibble(behind=lag(x),x,ahead=lead(x))

#case_when()
#-allows us to write a multi-condition statement without nesting lots of ifelse() statements.
iris %>% mutate(size=case_when(
  Sepal.Length<mean(Sepal.Length)~"s",
    Sepal.Length>mean(Sepal.Length)~"l",
TRUE~"m"
))

iris %>%
mutate(Species = case_when(
Species == "setosa" ~ toupper(Species),
TRUE ~ as.character(Species)
))
```

There are extra cumulative functions that consider all the rows
before the present one. We already have cumsum(), cummin(),
cummax() and a few others from base R. tidyverse gives us:
• cumall() for checking a condition has been TRUE in every
prior row
• cumany() for checking a condition has been TRUE in any
prior row
• cummean() for getting the mean of all prior values
The cumall() and cumany() are especially useful for saying
whether things happened in the past, along with lag(), as you
often want to be able to create such measures as candidate
columns when building predictive models.

```{r}
?cumsum
cumsum(1:10)

iris %>%
mutate(runagg=cumall(Sepal.Length<mean(Sepal.Length))
       )
```

## transmute
 
mutate() takes your data, applies the changes, and returns the updated dataset in its entirety. Occasionally, you might want to return only the columns you want to change. In such a situation,swap the word mutate for transmute.

```{r}


iris %>%
transmute(Sepal.Width=floor(Sepal.Width),
Species = case_when(
Species == "setosa" ~ toupper(Species),
TRUE ~ as.character(Species)
))
```

## Advanced data change

-Using mutate_all() you can apply a function to every column.
-If you want to change columns where they meet some sort of criteria within the data, you can provide a criteria to mutate_if().
-To change columns with certain name conditions use the
mutate_at() variant.
```{r}
iris %>%
mutate_all(as.character)

iris %>%
mutate_if(is.numeric, ~ . + rnorm(.))

iris %>%
mutate_at(vars(Sepal.Width:Petal.Width), ~ . + rnorm(.))

```


## Challenge 4
```{r}
#1
irisImperial<-iris %>% mutate(Species=toupper(Species),
  Sepal.Length=Sepal.Length/2.5,
  Sepal.Width=Sepal.Width/2.5,
  Petal.Width=Petal.Width/2.5,
  Petal.Length=Sepal.Length/2.5)


irisImperial

#2
library(ggplot2movies)
median(movies$length)
movies %>% mutate(length_median=length-median(length))
#3
irisImperial1<-iris %>% mutate_if(is.numeric,~./2.5) %>% mutate(Species=toupper(Species))
irisImperial1
```





```{r}
#revision
iris_imperial<-iris %>% mutate(Species=toupper(Species),
  Sepal.Length=Sepal.Length/2.5,
  Sepal.Width=Sepal.Width/2.5,
  Petal.Width=Petal.Width/2.5,
  Petal.Length=Sepal.Length/2.5)


iris_imperial
```



```{r}
fix(iris)
iris
iris$Sepal.Length[is.na(iris$Sepal.Length)]
sum(is.na(iris))
which(is.na(iris))
?which
?lag
mean(iris$Sepal.Length,na.rm = TRUE)
```

library
```{r}
library(tidyverse)
```























